### 2018-03-22 0_intro.tex
Cleanup wmm intro part. Set up named references. Fix too many commas error. Fix other citing errors.

The SC model is considered to be the strong memory model in the sence that it provides strong guarantees regarding the ordering and caused effect of memory operations. Different relaxations of this model lead to the class of \textit{weak memory models~(WMM)}.
%while preserving consistency
WMMs serve as set of guarantees made by designers of execution environment (hardware, programming language, compiler, database, operation system, etc.) to programmers on which behaviours of their concurrent code they may expect.

>>>>

The SC model is considered to be the strong memory model in the sence that it provides strong guarantees regarding the ordering and caused effect of memory operations. Different relaxations of this model lead to the class of \textit{weak memory models~(WMM)}.
%while preserving consistency
They specify how threads interact through shared memory, when a write becomes visible to other threads and what value a read can return. 
Therefore, WMMs serve as set of guarantees made by designers of execution environment (hardware, programming language, compiler, database, operation system, etc.) to programmers on which behaviours of their concurrent code they may expect.

### 2018-03-22 1_wmm.tex
Cleanup wmm intro part. Set up named references. Fix too many commas error. Fix other citing errors.

%"Semantics and verification under weak memory models have been the subject of study at least since 2007." [PORTHOS]
The \textit{weak memory model} is a set of predicative constraints on possible executions of a concurrent program. The study of formalisation the weak memory models for different architectures is being rapidly developed over last decade. Research of WMM aimes, firstly, to formalise the weak memory models and provide systematic, sound and complete formal approach of defining WMMs in order to be able to verify systems with respect to them.

%The generic framework of program analysis under weak memory models presented in~\ref{alglave2010shared} is discussed in Section~\ref{ch:wmm:event}.

Secondly, researchers work on extracting the formal hardware memory models from existing implementations of from their specifications, that are written in natural language and thus suffer from ambiguities and incompletenesses. Over last decade the memory models have been extracted for most mainstream multiprocessor architectures, such as x86-TSO (\textit{Total Store Order}) model for x86 architecture formalised in 2009~\cite{owens2009better}, SPARC-TSO for Sparc architecture~\cite{?!?!?!}, much more relaxed memory model ??? for Power and ARM architectures defined in ???~\cite{sarkar2011understanding}, Alpha (???). Moreover, in 2005 <who?Milner> started the work on developing the weak memory model for C++ language, which was introduced in C++11 (?) standard~\cite{???} (//todo: C11 MM ~\cite{batty2011mathematizing}). The memory model for Java that is based on the \textit{happens-before} principle was introduced in JVM??? in ??<year>.

Thirdly, important research direction targets the problem of verifying (or at least finding bugs in) existing software systems with respect to weak memory models. In this <area> the notable works <are on> defining the Linux kernel memory model <that is being actively developing these days~\cite{kernel_wmm_1}>. Distributed databases <also need the wmm, see transaction consistency~\cite{bailis2013highly}>

%wmm as a formal way to define guaranties that a hw, programming language, execution environment provide for programmers.
%considering wmm as a set of allowed behaviours, the latter wmms are the supersets
%wmm allows and disallows optimisations: partial sync of memory buffers, out-of-order execution (reordering), <more> => more behaviours that are unallowed in SC.
%question possible to answer with wmm: which behaviours (in addition to SC) are allowed? which new states are allowed? Consequently, correctness, absence of data races, deadlocks or portability issues, etc.

%One important question that rises naturally when talking about weak memory models is whether a program is portable from one processor architecture to another. This can be solved ...<ref to PORTHOS>

% more examples of what wmms allow to do?

@@

The semantics of a concurrent program is represented by the set of allowed executions.
An execution is uniquely defined by the set $\mathbb{X}$ of events have been executed in each thread (the \textit{control-flow} of a program), and the relations $\mathtt{rf}$ and $\mathtt{co}$ (\textit{data-flow} of a program)~\ref{alglave2010shared}. As it was shown in~\ref{wickerson2017automatically}, it is enough for memory models to constrain the executions independently instead of constraining the program as a whole.

@@

The Porthos tool uses the language \texttt{CAT}~\ref{alglave2016syntax} for defining weak memory models.

>>>>

%"Semantics and verification under weak memory models have been the subject of study at least since 2007." [PORTHOS]
The \textit{weak memory model} is a set of predicative constraints on possible executions of a concurrent program. The study of formalisation weak memory models for different architectures has been rapidly developed over last decade.

Firstly, the research of WMM aimes to formalise the weak memory models and provide systematic, sound and complete formal approach of defining WMMs in order to be able to verify systems with respect to them.
A generic framework of program analysis with respect to weak memory models was presented in 2010~\cite{alglave2010shared}. Similar approach is used \texttt{PORTHOS} as it is described
% in Section~\cite{ch:wmm:event}. 
further in current Chapter.

Secondly, researchers work on extracting the WMMs of hardware architectures from existing implementations of from their specifications, that are written in natural language and thus suffer from ambiguities and incompletenesses. Over last decade the memory models have been extracted for most mainstream multiprocessor architectures, such as x86-TSO (\textit{Total Store Order}) model for x86 architecture formalised in 2009~\cite{owens2009better}, much more relaxed memory model for Power and ARM architectures~\cite{alglave2009semantics,sarkar2011understanding}, for Alpha~\cite{?} and Sparc~\cite{?} memory models were defined in the specification. Moreover, most modern high-level programming languages rely on relaxed memory model. Thus, the memory model for Java that is based on the \textit{happens-before} principle~\cite{lamport1978time} was introduced in J2SE 5.0 in 2004~\cite{manson2005java}; the C++11 standard~\cite{?} has introduced the set of hardware-independent synchronisation fences and atomic operations, whenever the C++17 memory model is based on the relation \textit{strongly happens-before}.
%batty2011mathematizing

Thirdly, important research direction targets the problem of verifying (or at least finding bugs in) existing software systems with respect to weak memory models. Perhaps, the most notable work in this field is the project for defining the Linux kernel memory model, which is being actively developing these days~\cite{kernel1}.
%Distributed databases <also need the wmm, see transaction consistency~\cite{bailis2013highly}>

%experiments with kernel?

@@

The semantics of a concurrent program is represented by the set of allowed executions.
An execution is uniquely defined by the set $\mathbb{X}$ of events have been executed in each thread (the \textit{control-flow} of a program), and the relations $\mathtt{rf}$ and $\mathtt{co}$ (\textit{data-flow} of a program)~\cite{alglave2010shared}. As it was shown in~\cite{wickerson2017automatically}, it is enough for memory models to constrain the executions independently instead of constraining the program as a whole.

@@

The Porthos tool uses the language \texttt{CAT}~\cite{alglave2016syntax} for defining weak memory models.
